---
title: "project2-analysing stormdata"
output: html_document
---
This project is to explore the NOAA Storm Database and answer some basic questions about severe weather events.

Download data file from url and save it. Read the data. Load necessary packages.

```{r}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", destfile = "repdata-data-StormData.csv")

stormdata <- read.csv("repdata-data-StormData.csv")

suppressMessages(library(dplyr))
library(ggplot2)

```

Simplify and clean the data.
```{r}
sd_sub <- stormdata[, c(8, 23:28)]

sd_sub$EVTYPE[grep("WIND", sd_sub$EVTYPE)] <- "WIND"
sd_sub$EVTYPE[grep("HEAT", sd_sub$EVTYPE)] <- "HEAT"
sd_sub$EVTYPE[grep("SNOW", sd_sub$EVTYPE)] <- "SNOW"
sd_sub$EVTYPE[grep("TSTM", sd_sub$EVTYPE)] <- "THUNDERSTORM"
sd_sub$EVTYPE[grep("COLD", sd_sub$EVTYPE)] <- "COLD"
sd_sub$EVTYPE[grep("RAIN", sd_sub$EVTYPE)] <- "RAIN"
sd_sub$EVTYPE[grep("FLOOD", sd_sub$EVTYPE)] <- "FLOOD"
sd_sub$EVTYPE[grep("HURRICANE", sd_sub$EVTYPE)] <- "HURRICANE"

sd_sub$PROPDMGEXP <- as.character(sd_sub$PROPDMGEXP)
sd_sub$CROPDMGEXP <- as.character(sd_sub$CROPDMGEXP)
unique(sd_sub$PROPDMGEXP)
unique(sd_sub$CROPDMGEXP)
sub_PDEXP <- c("M","m","B","5","6","7","8")
sub_CDEXP <- c("M","m","B")

sub1 <- filter(sd_sub, FATALITIES >=1 | INJURIES >= 1 | PROPDMGEXP %in% sub_PDEXP | CROPDMGEXP %in% sub_CDEXP)

unique(sub1$PROPDMGEXP)
unique(sub1$CROPDMGEXP)

sub1$PROPDMGEXP[sub1$PROPDMGEXP == "B"] <- "1000000000"
sub1$PROPDMGEXP[sub1$PROPDMGEXP == "M"] <- "1000000"
sub1$PROPDMGEXP[sub1$PROPDMGEXP == "m"] <- "1000000"
sub1$PROPDMGEXP[sub1$PROPDMGEXP == "K"] <- "1000"
sub1$PROPDMGEXP[sub1$PROPDMGEXP == "H"] <- "100"
sub1$PROPDMGEXP[sub1$PROPDMGEXP == "5"] <- "100000"
sub1$PROPDMGEXP[sub1$PROPDMGEXP == "6"] <- "1000000"
sub1$PROPDMGEXP[sub1$PROPDMGEXP == "7"] <- "10000000"
sub1$PROPDMGEXP[sub1$PROPDMGEXP == "8"] <- "100000000"
sub1$PROPDMGEXP[sub1$PROPDMGEXP == "0"] <- "1"
sub1$PROPDMGEXP[sub1$PROPDMGEXP == ""] <- "1"
sub1$PROPDMGEXP[sub1$PROPDMGEXP == "-"] <- "1"

sub1$CROPDMGEXP[sub1$CROPDMGEXP == "B"] <- "1000000000"
sub1$CROPDMGEXP[sub1$CROPDMGEXP == "M"] <- "1000000"
sub1$CROPDMGEXP[sub1$CROPDMGEXP == "m"] <- "1000000"
sub1$CROPDMGEXP[sub1$CROPDMGEXP == "K"] <- "1000"
sub1$CROPDMGEXP[sub1$CROPDMGEXP == ""] <- "1"
sub1$PROPDMGEXP[sub1$PROPDMGEXP == "?"] <- "1"
sub1$PROPDMGEXP[sub1$PROPDMGEXP == "0"] <- "1"

sub1$PROPDMGEXP <- as.numeric(sub1$PROPDMGEXP)
sub1$CROPDMGEXP <- as.numeric(sub1$CROPDMGEXP)

sub2 <- sub1 %>% mutate(PROP = PROPDMG*PROPDMGEXP, CROP = CROPDMG*CROPDMGEXP)

```

1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
```{r}
summary_by_event <- sub2 %>% group_by(EVTYPE) %>% 
    summarize_each(funs(sum), FATALITIES, INJURIES, PROP, CROP)

totalfatal <- summary_by_event %>% select(EVTYPE, FATALITIES) %>%
  arrange(desc(FATALITIES)) 
top10fatal <- totalfatal[1:10,]
top10fatal

ggplot(top10fatal, aes(x=reorder(EVTYPE, -FATALITIES), y=FATALITIES)) + geom_bar(stat = "identity", fill = "blue") + labs(y="Total Fatalities", x="Event") + theme(axis.text.x = element_text(angle = 90)) + ggtitle("Events with the top 10 fatalies")

totalinj <- summary_by_event %>% select(EVTYPE, INJURIES) %>%
  arrange(desc(INJURIES)) 

top10inj <- totalinj[1:10,]
top10inj

ggplot(top10inj, aes(x=reorder(EVTYPE, -INJURIES), y=INJURIES)) + geom_bar(stat = "identity", fill = "blue") + labs(y="Total Injuries", x="Event") + theme(axis.text.x = element_text(angle = 90)) + ggtitle("Events with the top 10 injuries")

```


2. Across the United States, which types of events have the greatest economic consequences?
```{r}
totalpropdmg <- summary_by_event %>% select(EVTYPE, PROP) %>%
  arrange(desc(PROP)) 
top10propdmg <- totalpropdmg[1:10,]
top10propdmg

ggplot(top10propdmg, aes(x=reorder(EVTYPE, -PROP), y=PROP)) + geom_bar(stat = "identity", fill = "blue") + labs(y="Total Property Damages", x="Event") + theme(axis.text.x = element_text(angle = 90)) + ggtitle("Events with the top 10 property damages")

totalcropdmg <- summary_by_event %>% select(EVTYPE, CROP) %>%
  arrange(desc(CROP)) 
top10cropdmg <- totalcropdmg[1:10,]
top10cropdmg

ggplot(top10cropdmg, aes(x=reorder(EVTYPE, -CROP), y=CROP)) + geom_bar(stat = "identity", fill = "blue") + labs(y="Total Crop Damages", x="Event") + theme(axis.text.x = element_text(angle = 90)) + ggtitle("Events with the top 10 crop damages")

```
